---
title: "Bacteroides vulgatus refseq/external genomes"
author: "PennCHOP Microbiome Program (Chunyu Zhao)"
date: "11/22/2017"
output: 
  pdf_document:
    includes:
      in_header: header.tex 
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  tidy=FALSE,
  cache=FALSE,
  echo=FALSE,
  message = FALSE,
  warning = FALSE,
  dpi=100,
  fig.width = 10,
  fig.height = 10,
  fig.align = "center",
  cache.lazy = FALSE,
  dev=c("png", "pdf"),
  warning=FALSE)
```

```{r}
library(dplyr)
library(ggplot2)
library(qiimer)
library(readr)
library(tidyr)
library(magrittr)
library(reshape2)
library(pheatmap)
library(pander)
library(broom)

library(subfunc)
library(grid)
library(stringr)
library(vegan)
library(ape)
library(forcats)
library(lubridate)
library(magrittr)

library(colorRamps)
library(RColorBrewer)

library(ggbeeswarm)

library(nlme)

source("helper.R")

require(Biostrings)
library(Rsamtools)

```

# Read in sunbeam pipeline and metadata

In this report, we looked at the `meconium` samples from IGRAM run1, and identified the strain of Bacteroide vulgatus from the annotation of de novo assembled contigs.

```{r sample_sheet_import, echo=FALSE}
mapping_file_name <- "mapping_file.txt"

s <- read.delim("igram_meconium_sampleinfo.txt", sep="\t", stringsAsFactors = FALSE) 
  
s %<>% mutate(SampleID = ifelse(is.na(SampleID), as.character(sample_code), as.character(SampleID)))

color_by <- "SampleType"
shape_by <- ""
```

```{r}
preprocess <- read.delim(file.path("sunbeam_output","preprocess_summary.tsv")) %>%
  mutate(Samples = sub(".json", "", Samples)) %>%
  mutate(low_quality = (fwd_only + rev_only + dropped) / input) %>%
  mutate(human = true / input) %>%
  mutate(non_human = false / input) 

s <- merge(s, preprocess, by.x="SampleID", by.y = "Samples")
```

# Read in metaphlan2 results

```{r}
#- 20170912
remove_rank_prefix <- function (x) {
  gsub("[kpcofgs]__", "", x, perl=TRUE)
}

replace_underscores <- function (x) {
  gsub("_+", " ", x, perl=TRUE)
}

read_metaphlan2_table <- function (filepath, sample_prefix="PCMP_") {
  df <- read.delim(filepath, comment.char="#")
  props <- as.matrix(df[,-1])
  colnames(props) <- sub(sample_prefix, "", colnames(props), fixed = TRUE)
  rownames(props) <- replace_underscores(remove_rank_prefix(df[,1]))
  props
}

props <- read_metaphlan2_table("sunbeam_output/taxonomic_assignments.tsv")

#- we need to clean up the names (20170912)
filenames <- colnames(props)
filenames = sub("-","_", filenames)
colnames(props) <- filenames


name1 = colnames(props)
name2 = s$SampleID
if(dim(props)[2]!= dim(s)[1]){
  # filter out control samples without metaphlan2 result
  missing_names = setdiff(name2,name1)
  s <- s %>% filter(! SampleID %in% missing_names)

  #- check for SampleID mismatch
  missing_names = setdiff(name1, name2)
  props <- props[,! name1 %in% missing_names]
}

adf <- split_assignments(rownames(props), split = "\\|")
a <- simplify_assignments(adf, rank2 = "Species") 

rownames(props) <- a

genus_props <- props
```

## Dominance

```{r dominance, fig.width=8}
top_df <- genus_props%>%
  melt(c("Taxon", "SampleID"), value.name="Proportion") %>% 
  merge(s, by= "SampleID") %>% 
  group_by(SampleID) %>%
  filter(Proportion == max(Proportion)) %>%
  dplyr::slice(1) %>%
  ungroup() %>%
  select(SampleID, Taxon, Proportion) %>%
  dplyr::rename(Dominance = Taxon) %>%
  dplyr::rename(DominanceProp = Proportion)

s %<>% merge(top_df, by="SampleID") %>%
  mutate(Dominance = as.factor(Dominance))

fig <- s %>% 
  group_by(Dominance) %>%
  summarise(Count= n()) %>%  
  mutate(Count = Count / dim(s)[1]) %>%
  arrange(desc(Count)) %>%
  mutate(Dominance = factor(Dominance, levels = .$Dominance)) %>%
  ggplot(aes(x=factor(1), y=Count, fill=Dominance)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +
  theme_minimal() +
  ggtitle("Dominace for all the samples")
```

```{r}
genus_props_df <- genus_props %>%
  melt(c("Taxon", "SampleID"), value.name = "Proportion")
```

# Bacteroides vulgatus

The Escherichia coli genome was found to be a circular DNA molecule 4.7 million base pairs in length, containing 4288 annotated protein-coding genes (organized into 2584 operons), seven ribosomal RNA (rRNA) operons, and 86 transfer RNA (tRNA) genes.

- also need an outgroup**

- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3147508/

## Sample Selections

- write 20171122_Bvulgatus.txt
- write 20171122_SampleID_Bvulgatus.txt

```{r}
props_toadd <- genus_props_df %>% filter(Taxon %in% "Bacteroidetes Bacteroides vulgatus") %>% droplevels()

s_toTest <- s %>%
  select(SampleID, NonHostReads, both_kept, false, Dominance) %>%
  merge(props_toadd) %>%
  mutate(estNumReads=Proportion * false) %>%
  mutate(estCoverage=estNumReads / 4700000 ) %>%
  arrange(desc(estCoverage))

s_toTest %>% filter(estCoverage > 0 ) %>%
  write.table("20171122_Bvulgatus.txt", sep="\t", quote=FALSE, row.names=FALSE)

s_toTest %>%
  filter(estCoverage > 4 ) %>%
  select(SampleID) %>%
  write.table("20171122_SampleID_Bvulgatus.txt", sep="\t", quote=FALSE, row.names=FALSE, col.names=FALSE)
```


## Assembly summary

- read in 20171122_Bvulgatus/assembly_summary.txt
- write 20171122_Bvulgatuss_accession.txt

```{r}
genome_info <- read.delim("20171122_Bvulgatus/assembly_summary.txt", sep='\t', stringsAsFactors = FALSE) 

genome_info %<>% filter(! infraspecific_name %in% "")

genome_info %<>% mutate(genome_name = paste("B vulgatus",infraspecific_name, sep=" "))

genome_info %>% select(assembly_accession) %>%
   write.table("20171122_Bvulgatuss_accession.txt", quote=FALSE, row.names=FALSE, col.names=FALSE)
```

## External genome

- need to double check for `.` in the name

- write 20171122_vulgatus-external-genomes.txt

```{r}
library(lubridate)

temp <- genome_info %>%
  select(genome_name, assembly_accession) %>%
  mutate(assembly_accession = gsub("\\.\\d", "", assembly_accession)) %>%
  mutate(contigs_db_path = paste0(assembly_accession, ".db")) %>%
  dplyr::rename(name = genome_name) %>%
  mutate(name=gsub(" ","_", name)) %>%
  mutate(name=gsub("-","_", name)) %>%
  mutate(name=gsub(":","_", name)) %>%
  mutate(name=gsub("substr\\._","", name)) %>%
  mutate(name=gsub("str\\._","", name)) %>%
  mutate(name=gsub("=","_", name)) %>%
  mutate(name=gsub("\\(","", name)) %>%
  mutate(name=gsub("\\)","", name)) %>%
  select(-assembly_accession)

temp %>% write.table("20171122_vulgatus-external-genomes.txt", quote=FALSE, row.names=FALSE, sep="\t")
```

```{bash, eval=FALSE}
#1. convert fasta files into 
for i in *.fa
do
	anvi-script-FASTA-to-contigs-db $i
done
```

