---
title: "Generate Panphlan related figures and analysis for the Paper"
author: "Chunyu Zhao"
date: \today
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  tidy=FALSE,
  cache=FALSE,
  echo=FALSE,
  message = FALSE,
  warning = FALSE,
  dpi=100,
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  cache.lazy = FALSE,
  dev=c("png", "pdf"),
  warning=FALSE)
```

```{r load scripts, echo=FALSE, message=FALSE, warning=FALSE}
library(qiimer)
library(ape)
library(vegan)
library(ggplot2)
library(dplyr)
library(reshape2)
library(kylemisc)
library(tidyr)
library(broom)
library(pheatmap)

library(subfunc)
library(forcats)
library(lubridate)
library(magrittr)
library(pander)
library(ggbeeswarm)
library(tidyr)

library(RColorBrewer)
library(readr)
```

```{r}
source("helper.R")
```

# pangenome

## read in pan matrix

- read in result_gene_presence_absence_20180721.csv
- read in 20180714_ecoli_tree_tip_order.txt
- read in 20180604_tree_dd.txt

```{r}
read_pan_matrix <- function(filepath, commented = TRUE, metadata = TRUE) {
  f <- file(filepath, "rt")
  header_line <- readLines(f, n = 1)
  
  if (commented){
    header_line <- readLines(f, n = 1)
  }
  
  col_names <- strsplit(header_line, "\t")[[1]]
  col_classes <- rep("numeric", times = length(col_names))
  col_classes[1] <- "character"
  
  full_otu_table <- read.table(f, col.names = col_names, colClasses = col_classes,
                               sep = "\t", quote = "", as.is = TRUE, header = FALSE)
  close(f)
  
  data_cols <- 2:length(col_names)
  
  sample_ids <- col_names[data_cols]
  otu_ids <- as.character(full_otu_table[, 1])
  counts <- as.matrix(full_otu_table[, data_cols])
  rownames(counts) <- otu_ids
  
  list(sample_ids = sample_ids, otu_ids = otu_ids, counts = counts)
}

o <- read_pan_matrix("result_gene_presence_absence_20180721.csv", commented=FALSE)
cts <- o$counts
props <- sweep(cts, 2, colSums(cts), "/")

strain_names <- colnames(cts)
gene_names <- rownames(cts) 

tree.tip.order <- read_delim("20180714_ecoli_tree_tip_order.txt", delim="\t", col_names = F) 
tree.tip.order <- tree.tip.order$X1

dd <- read_delim("20180604_tree_dd.txt", delim="\t") %>%
  select(accession, taxon, node_lable) %>%
  mutate(taxon = factor(taxon, levels= tree.tip.order))
```

- read in 20171129_E.coli_strains_summary.tsv

```{r}
strains <- read.table("20171129_E.coli_strains_summary.tsv", sep="\t", header = TRUE)

strains %<>% select(organism_name, pathovar, assembly_accession, taxid, species_taxid, phylo_group) %>%
  separate(assembly_accession, into="accession",extra="drop",sep="\\.") 

strains_df <- data.frame(strain_names = strain_names) %>%
  mutate(accession = gsub("REF_", "", strain_names)) %>% 
  left_join(strains, by=c("accession")) %>% #, by=c("strain_names"="accession")
  mutate(node_lable = ifelse(is.na(pathovar), as.character(accession), as.character(pathovar)))

strains_df %<>% 
  mutate(phylo_group = ifelse(grepl("Shigella", organism_name), "Shigella", as.character(phylo_group))) %>%
  mutate(phylo_group = ifelse(grepl("^s", strain_names), "PCMP", as.character(phylo_group))) %>% 
  mutate(phylo_group = ifelse(is.na(phylo_group), "Unknown", as.character(phylo_group))) %>% 
  mutate(phylo_group = factor(phylo_group, levels=c("A", "B1", "B2", "C-I", "D1", "D2", "E", "PCMP", "Shigella", "Unknown"))) %>%
  mutate(pathovar = ifelse(grepl("^s", accession) & is.na(pathovar), "PCMP", as.character(pathovar))) %>%
  mutate(pathovar = factor(pathovar, levels=c("Commensal","Laboratory adapted","EAEC","EPEC","ETEC","EHEC","ExPEC/UPEC","ExPEC/avian","PCMP", "NA"))) 
```

## heatmap 

```{r}
color = saturated_rainbow(101)
breaks = c(0, 1e-10, seq(0.001, 1, length.out = 100))

heatmap_cts <- cts[,strains_df$strain_names]
rows_to_keep <- filter_low_coverage(heatmap_cts, perc_cutoff=0.8)
top_genera <- rownames(heatmap_cts[rows_to_keep, ])

grps <- c("phylo_group","pathovar")
annc <- strains_df[,grps] %>% as.data.frame()
rownames(annc) <- strains_df$strain_names
colnames(annc) <- grps
```

## 20180714_Fig2A_pan_heatmap.pdf

```{r, eval=FALSE}
color = colorRampPalette(brewer.pal(n = 9, name = "RdPu"))(9)
color <- c("white", color[c(4)])
breaks = seq(0, 1, length.out = 3)

row.order <- hclust(dist(cts.toplot))$order
col.order <- levels(dd$accession)
dat_new <- cts.toplot[row.order, col.order] # re-order matrix accoring to clustering

# melt to dataframe
df_molten_dat <- reshape2::melt(as.matrix(dat_new)) # reshape into dataframe
names(df_molten_dat)[c(1:3)] <- c("gene_family", "genome", "counts")

df_molten_dat %>% 
  mutate(counts = as.factor(counts)) %>%
  ggplot(aes(x = gene_family, y = genome, fill = counts)) + 
  geom_tile() +
  scale_fill_manual(values = color) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks=element_blank()) +
  ggsave("20180714_Fig2A_pan_heatmap.pdf", width=18, height=8, useDingbats=FALSE)
```


## 20180604_pcoa.pdf

Here, we use Jaccard distance to compare samples based on shared species membership.  Plots are described above.

```{r}
jc <- dist(t(cts), method = "binary")
pc <- pcoa(jc)

matrix_to_sample_df <- function (x) {
  df <- data.frame(SampleID=rownames(x), stringsAsFactors = FALSE)
  cbind(df, x)
}

pc_df <- matrix_to_sample_df(pc$vectors[,1:3])

pc_df %<>% merge(strains_df, by.x="SampleID", by.y="strain_names")

pc_pct <- round(pc$values$Relative_eig * 100)
```


```{r}
library(ggsci)

mycolor <- pal_npg(palette = c("nrc"), alpha = 1)(10)

pc_df %<>% mutate(isPCMP = ifelse(phylo_group %in% "PCMP", TRUE, FALSE)) %>%
  mutate(isPCMP = as.factor(isPCMP))

fig <- pc_df %>% 
  ggplot(aes(x=Axis.1, y=Axis.2, color=phylo_group, shape=isPCMP)) +
  geom_point(aes_string(x="Axis.1", y="Axis.2")) +
  theme_bw() +
  labs(x = paste0("PCoA axis 1 (", pc_pct[1], "%)"), y = paste0("PCoA axis 2 (", pc_pct[2], "%)")) +
  coord_equal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_npg() +
  scale_shape_manual(values = c(16, 8), guide=FALSE) 
fig + ggsave("20180604_pcoa.pdf", width=3.5, height=3, useDingbats=FALSE)


pc_df %>%
  ggplot(aes(x=Axis.1, y=Axis.2, color=pathovar, shape=isPCMP)) +
  geom_point(aes_string(x="Axis.1", y="Axis.2")) +
  theme_bw() +
  labs(x = paste0("PCoA axis 1 (", pc_pct[1], "%)"), y = paste0("PCoA axis 2 (", pc_pct[2], "%)")) +
  coord_equal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_shape_manual(values = c(16, 8), guide=FALSE) 
```

## 20181029 Hierarchical clustering based on Jaccard distance

```{r,fig.width=8, fig.height=8}
setdiff(sub("REF_", "", strains_df$strain_names), dd$accession)
print("the missing strains from the panphaln analysis")
setdiff(dd$accession, sub("REF_", "", strains_df$strain_names))

old.names <- colnames(cts)
DF <- strains_df %>%
  mutate(organism_name = ifelse(is.na(organism_name), as.character(strain_names), as.character(organism_name))) %>% 
  filter(strain_names %in% old.names) %>%
  select(strain_names, organism_name) %>%
  mutate(strain_names = factor(strain_names, levels=old.names)) %>% 
  arrange(strain_names) %>%
  mutate(organism_name = factor(organism_name, levels=.$organism_name))
new.names <- levels(DF$organism_name)
colnames(cts) <- new.names

is.Shigella <- grepl("Shigella", colnames(cts))
cts <- cts[,!is.Shigella, drop=FALSE]


jc <- dist(t(cts), method = "binary")

jc_upgma <- hclust(jc, method = "complete")
pdf("20181029_pangenome_jc_cluster.pdf", width=5, height = 4)
par(cex=0.5)
plot(
  jc_upgma, hang=-1,
  ylab = "Jaccard distance",
  xlab="Hierarchical clsutering",
  sub="Complete-linkage method (UPGMA)")
par(cex=1)


jc_upgma <- hclust(jc, method = "average")
par(cex=0.5)
plot(
  jc_upgma, hang=-1,
  ylab = "Jaccard distance",
  xlab="Hierarchical clsutering",
  sub="Average-linkage method (UPGMA)")
par(cex=1)


jc_upgma <- hclust(jc, method = "single")
par(cex=0.5)
plot(
  jc_upgma, hang=-1,
  ylab = "Jaccard distance",
  xlab="Hierarchical clsutering",
  sub="Single-linkage method (UPGMA)")
par(cex=1)


jc_upgma <- hclust(jc, method = "ward.D")
par(cex=0.5)
plot(
  jc_upgma, hang=-1,
  ylab = "Jaccard distance",
  xlab="Hierarchical clsutering",
  sub="Ward-linkage method (UPGMA)")
par(cex=1)

dev.off()
```


\newpage
## 20180602 Fisher's exact test

- We compare the differential abundance of sparsely-sampled (rare) features using Fisher's exact test. Fisher's exact test models the sampling process according to a hypergeometric distribution (sampling without replacement)
- Plot fraction of samples in each group for which the taxon is present.
  
* filter out gene clusters show up in only two samples, or not show up in only two samples

- write 20180723_Fig2D_gene_function.txt
- write 20180604_fisher.pdf

```{r, fig.width=5, fig.height=5}
#https://github.com/road2stat/ggsci
props_toTest <- props
s_toTest <- strains_df %>% select(accession) %>% 
  mutate(Group = ifelse(grepl("^s",accession),"PCMP", "non-PCMP"))

genus_present <- apply(props_toTest, 2, function (x) x > 0) %>%
  melt(c("Gene", "SampleID"), value.name="Present") %>%
  mutate(SampleID = sub("REF_", "", SampleID)) %>%
  merge(s_toTest, by.x="SampleID", by.y = "accession")

genus_present %>%
  count(Group)

low_lm <- 2
high_lm <- ncol(props_toTest)-2
genus_present %<>%
  group_by(Gene) %>%
  filter(sum(Present) < high_lm & sum(Present) > low_lm) %>%
  ungroup()

ret <- genus_present %>%
  group_by(Gene) %>%
  filter(sum(Present) > low_lm, sum(Present) < high_lm) %>%
  do(tidy(fisher.test(with(., table(Present, Group))))) %>%
  ungroup() %>%
  arrange(p.value) %>%
  mutate(fdr_corrected = p.adjust(p.value, method="fdr")) %>%
  filter(p.value <= 0.05) 

ret.sig <- ret %>% 
  arrange(fdr_corrected) %>%
  mutate(isSig = ifelse(fdr_corrected <= 0.01, "*", "")) %>%
  slice(1:10)

pander(ret.sig %>% select(-one_of("method", "alternative")))

genes.toplot.df <- data.frame(Gene = as.character(ret.sig$Gene)) %>%
  droplevels() %>%
  mutate(Gene = as.character(Gene)) %>%
  mutate(FacetGroup = ifelse(row_number() <= 5, "A", "B"))

## for the new version of panphaln, we can actually read in the function
annot <- read_delim("ecoli.database_20180722/panphlan_ecoli-20180720_annotations.csv", delim="\t")

annot %>% 
  filter(Gene_family %in% genes.toplot.df$Gene) %>% 
  select(Gene_family, Centroid_gene_ID, Product) %>%
  write.table("20180723_Fig2D_gene_function.txt", sep="\t", quote = F, row.names = F)

fig <- genus_present %>% 
  inner_join(ret.sig, "Gene") %>%
  merge(genes.toplot.df, by="Gene") %>% 
  mutate(Gene = paste(Gene, isSig, sep="")) %>% 
  mutate(Gene = as.factor(Gene)) %>%
  group_by(Gene, Group) %>%
  mutate(frac_present = sum(Present) / n()) %>% 
  ungroup() %>% 
  mutate(Gene = reorder(droplevels(Gene), -p.value)) %>% 
  ggplot() +
  geom_bar(aes(x=Gene, y=frac_present, fill=Group), position="dodge", stat="identity") +
  scale_fill_brewer(palette="Paired", guide=guide_legend(reverse=T)) +
  theme_bw() +
  labs(y="Fraction of samples where taxon is present", fill="", x="") +
  coord_flip() +
  facet_wrap(~FacetGroup, scales = "free") +
  theme(strip.background = element_blank(),strip.text.x = element_blank())

fig + ggsave("20180604_fisher.pdf", width=7, height=3, useDingbats=FALSE)
```
